---
title: "One Quantitative Variable Analysis Lecture Workbook"
format: html
editor: visual
---

# Specify the Elements of the Research Question

1.  Population of Interest\
2.  Variable of Interest\
3.  Characteristic of the Variable to be Analyzed\
4.  General Analytical Method

# Specify the Research Hypothesis

5.  The Research Hypothesis is:

# Specify the Statistical Hypotheses

6.  $H_A$:
7.  $H_0$:

# Descriptive Evidence Gathering Techniques

8.  Calculate the Sample Mean

    ```{r}
    ## Step 1: Read in the Data ##

    tiktok <- readxl::read_xlsx("TikTok Usage.xlsx")

    ## Step 2a: Calculate the Sample Mean ##

    mean(tiktok$`Daily Time`)

    ```

9.  Calculate the Sample Median

    ```{r}
    ## Step 2b: Calculate Sample Median ##

    median(tiktok$`Daily Time`)
    ```

10. Do these two values provide support for the research hypothesis?\

11. Calculate the sample standard deviation

```{r}
## Step 3: Calculate Sample Standard Deviation ##

sd(tiktok$`Daily Time`)
```

12. Provide an interpretation of the sample standard deviation\

13. Calculate Q1 and Q3

    ```{r}
    ## Step 4: Calculate Q1 & Q3 ##

    quantile(tiktok$`Daily Time`,probs=c(0.25,0.75))
    ```

14. Do Q1 and Q3 provide support for the research hypothesis?\

15. Calculate the appropriate test statistic and its associated p-value

16. 
